---
title: "TCSI: Backcast treelists from Landis summary data"
author: "Henry Rodman"
date: "August 3, 2020"
output: pdf_document
params:
  treelist_dir: NULL
  landis_dir: NULL
---

This is the Landis backcaster. It takes Landis output (biomass in grams per square meter of species x age bins) and converts back into tph by species and diameter bins.

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(dtplyr)

theme_set(theme_bw())
attach(params)
```

Read the landis files
```{r landis_import}
# these are all of the landis csv files derived from the 90m basemap treelists
all_landis_files <- list.files(landis_dir, full.names = TRUE)

# save one for testing the matching procedure
test_landis_file <- sample(all_landis_files, 1)

# use the rest for building lookup table
training_landis_files <- setdiff(
  all_landis_files,
  test_landis_file
)

landis_raw <- do.call(rbind, lapply(training_landis_files, data.table::fread))

lookup <- process_landis(landis_raw)

# import the other one for evaluating matching
new_data <- process_landis(
  data.table::fread(test_landis_file)
)

```

Backcast the Landis output into treelists
```{r backcast}
backcasted <- backcast_landis_to_treelists(
  new_data = new_data,
  lookup = lookup,
  n_clusters = 50,
  treelist_dir = treelist_dir
)

```

### Diagnostics
The `backcasted` object also contains two tables that are useful for generating diagnostics for the accuracy of the matching process. Generally speaking, the pixel matching process will do a better job matching the values of the species and age classes with the largest share of biomass for each pixel.

#### Total biomass

```{r total_biomass_plot, echo = FALSE}
backcasted$comp_frame %>%
  filter(attr == "aboveground_biomass_g_per_m2") %>%
  ggplot(aes(x = original, y = matched)) +
  geom_point(alpha = 1 / 10) +
  geom_abline() +
  labs(
    title = "total biomass (grams / sq. m)",
    x = "original estimate",
    y = "estimate from matched pixel"
  )
```

```{r total_biomass_table, echo = FALSE}
backcasted$comp_stats %>%
  filter(attr == "aboveground_biomass_g_per_m2") %>%
  transmute(
    attribute = attr,
    `original mean` = mean_original,
    `matched mean` = mean_matched,
    RMSE = rmse,
    `RMSE %` = rmse_pct
  ) %>%
  pander::pander()

```

#### Biomass by species

```{r species_biomass_plot, echo = FALSE}
backcasted$comp_frame %>%
  filter(
    attr %in% attr(lookup, "spp_vars")
  ) %>%
  ggplot(aes(x = original, y = matched)) +
  geom_point(alpha = 1 / 10) +
  geom_abline(color = "blue") +
  facet_wrap(~ attr) +
  labs(
    title = "species biomass (grams / sq. m)",
    x = "original estimate",
    y = "estimate from matched pixel"
  )
```

```{r species_biomass_table, echo = FALSE}
backcasted$comp_stats %>%
  filter(attr %in% attr(lookup, "spp_vars")) %>%
  arrange(desc(mean_original)) %>%
  transmute(
    species = attr,
    `original mean` = mean_original,
    `matched mean` = mean_matched,
    RMSE = rmse,
    `RMSE %` = rmse_pct
  ) %>%
  pander::pander()

```


#### Biomass by age class

```{r age_biomass_plot, echo = FALSE}

backcasted$comp_frame %>%
  filter(
    attr %in% attr(lookup, "age_vars")
  ) %>%
  ggplot(aes(x = original, y = matched)) +
  geom_point(alpha = 1 / 10) +
  geom_abline(color = "blue") +
  facet_wrap(~ attr) +
  labs(
    title = "age class biomass (grams / sq. m)",
    x = "original estimate",
    y = "estimate from matched pixel"
  )
```

```{r age_biomass_table, echo = FALSE}
backcasted$comp_stats %>%
  filter(attr %in% attr(lookup, "age_vars")) %>%
  mutate(
    age_class = as.numeric(gsub("age_", "", attr))
  ) %>%
  arrange(age_class) %>%
  transmute(
    `age class` = paste(age_class, age_class + 10, sep = " - "),
    `original mean` = mean_original,
    `matched mean` = mean_matched,
    RMSE = rmse,
    `RMSE %` = rmse_pct
  ) %>%
  pander::pander()

```
